---
title: "Getting started with documenteR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with documenteR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(documenteR)
options(documenteR.quiet = TRUE)
```

## The problem

A typical analysis script ends with a block like this:

```r
ggsave("outputs/plots/cars.png", plt_cars, width = 16, height = 12, units = "cm")
ggsave("outputs/plots/speed.png", plt_speed, width = 8, height = 8, units = "cm")
write.csv(data_cars, "outputs/data/cars.csv")
saveRDS(model, "outputs/model.rds")
```

Three months later nobody remembers what `speed.png` shows, where the data
came from, or which version of the script produced it. And when the numbers
change, every path and size has to be revisited.

`documenteR` inverts this. You mark each output *where it is created*, with
its documentation attached, and export everything in one call.

## Initialise

```{r}
out <- init_outputs(default = FALSE)
out
```

In a real script you would call `init_outputs()` with no arguments, which
creates `list_outputs` in your workspace and lets every later call omit
`output_list`. Here we keep the collection in a variable so the vignette
stays self-contained.

## Append as you go

### Data

```{r}
data_cars <- cars

append_data(
  data_cars,
  output_list = out,
  desc   = "Speed and stopping distance of cars in the 1920s",
  source = c(datasets = "R built-in dataset"),
  labels = c(speed = "Speed (mph)", dist = "Stopping distance (ft)"),
  format = c(".csv", ".rds")
)
```

`labels` feeds the codebook that ships with the export. If your data came
from SPSS or Stata via `haven`, the labels already stored on each column are
picked up automatically.

### Plots

Any plotting system works. A `ggplot2` object:

```{r eval = requireNamespace("ggplot2", quietly = TRUE)}
plt_cars <- ggplot2::ggplot(cars, ggplot2::aes(speed, dist)) +
  ggplot2::geom_point()

append_plot(
  plt_cars,
  output_list = out,
  desc = "Braking distance rises faster than speed",
  vars = c(x = "Speed (mph)", y = "Distance (ft)"),
  type = "m_wide"
)
```

A base-R plot, which cannot be captured as an object, is passed as a function
of no arguments:

```{r}
plt_hist <- function() {
  hist(cars$speed, col = "steelblue", border = "white",
       main = "Distribution of speeds", xlab = "Speed (mph)")
}

append_plot(plt_hist, output_list = out,
            desc = "Speeds are roughly symmetric", type = "s_wide")
```

`lattice` objects, `grid` grobs, `gtable`s, `patchwork` compositions and
recorded base plots are all accepted too. To support something else, define a
`dr_draw()` method for its class.

#### Sizes

`type` selects a preset from `gpars$sizes$type`. The defaults are built for
an A4 page with 2.5 cm margins, so `16` cm is the usable text width:

```{r}
names(out$gpars$sizes$type)
out$gpars$sizes$type$m_wide
```

`H`, `W` and `units` override a preset. To check a choice before running a
full export, use `preview_plot()`, which writes the plot at exactly the
geometry `save_outputs()` would use and opens it:

```r
preview_plot(plt_cars, type = "m_wide")
```

### Tables

`append_table()` is `append_data()` without the codebook. The distinction is
intent: `data` is the analysis-ready dataset you want others to reuse,
`tables` are presentation tables.

```{r}
tab_speed <- aggregate(dist ~ speed > 15, data = cars, FUN = mean)
names(tab_speed) <- c("fast", "mean_dist")

append_table(tab_speed, output_list = out,
             desc = "Mean stopping distance by speed group")
```

### In-text statistics

Numbers quoted in a report are the outputs most likely to go stale, and the
hardest to trace. `append_stats()` records the value, how it was computed,
and the sentence it appears in:

```{r}
stat_mean_speed <- mean(cars$speed)

append_stats(
  stat_mean_speed,
  output_list = out,
  type        = "float",
  stat_round  = 1,
  formula     = "sum(speed)/n",
  vars        = c(speed = "Speed of each car", n = "Number of cars"),
  text        = "Cars travelled at * mph on average.",
  section     = "2.1"
)
```

All appended statistics land in one table. `formula_text` is the formula with
the variable labels substituted in, and `text_fill` is the sentence with the
formatted number in place of `*` — ready to paste into a report:

```{r}
stats <- documenteR:::build_stats_table(out$stats, out$gpars)
stats[, c("name", "stat", "formula_text", "text_fill")]
```

## Review before exporting

Printing the collection shows what it holds, with the file name each output
will get, so it doubles as a preview of the export:

```{r}
out
```

The last line reports documentation coverage, which is the quickest way to
notice the plot nobody described. `print(out, detail = TRUE)` lists every
output and adds the description column.

`summary()` gives the full inventory as a data frame:

```{r}
summary(out)
```

Because it is an ordinary data frame underneath, you can ask it the obvious
question:

```{r}
s <- summary(out)
s$name[!s$has_desc]
```

And to see everything recorded about one output, print it:

```{r}
out$plots$plt_hist
```

## Export

```{r}
output_dir <- file.path(tempdir(), "vignette_outputs")

res <- save_outputs(
  output_dir    = output_dir,
  output_list   = out,
  project_title = "Cars example",
  version_notes = "First upload"
)

res$version
```

What was written:

```{r}
list.files(res$path, recursive = TRUE)
```

Before committing to an export, `dry_run = TRUE` reports exactly what would
be written and writes nothing.

### The README

Each output folder gets a plain-text README built from the documentation you
attached:

```{r comment = ""}
cat(get_output_README(out, "data"))
```

### The codebook

```{r}
utils::read.csv(file.path(res$path, "02_data", "00_codebook.csv"))
```

### The catalogue

`03_plots/00_01_catalogue.pdf` has one page per plot: its documentation
typeset at the top, the plot below at its true physical size. Every page is
the same size and orientation, and aspect ratios are preserved — a plot too
large for the page is scaled down proportionally, and the footer says by how
much. Configure it with `dr_catalogue_pars()`.

### The manifest and the log

The manifest records every file with the object it came from, its
documentation, its size and an MD5 checksum:

```{r}
man <- dr_manifest(output_dir)
man[, c("type", "name", "file", "bytes")]
```

`dr_verify()` re-computes those checksums, which is worth running after
copying an export across a network share. `dr_log()` shows what the export
recorded: messages and warnings from other packages are captured there
instead of being printed, so the progress report stays readable, and a single
output that cannot be written is marked `failed` in the manifest while the
rest of the export continues.

## Versions

Each export adds a version. Minor by default, major on request:

```{r}
save_outputs(output_dir, out, version_notes = "Fixed a label")
save_outputs(output_dir, out, new_version = TRUE, version_notes = "New method")

dr_versions(output_dir)[, c("version", "notes")]
```

`update_latest_version = TRUE` overwrites the most recent version folder
instead of adding one — for fixing a typo, not for new results.

## Configuring defaults

Everything adjustable lives in `gpars`. Start from `dr_gpars()`, which merges
your overrides into the defaults recursively, so you only name what you
change:

```{r}
out2 <- init_outputs(
  default = FALSE,
  gpars = dr_gpars(
    formats   = list(plots = c(".png", ".svg")),
    csv_bom   = TRUE,
    catalogue = dr_catalogue_pars(paper = "a3", orientation = "landscape")
  )
)
out2$gpars$formats$plots
out2$gpars$formats$data   # untouched
```

You can also change parameters after the fact, including adding your own
size presets:

```{r}
out2$gpars$sizes$type$wide_banner <- list(units = "cm", H = 6, W = 24)
names(out2$gpars$sizes$type)
```

`csv_bom = TRUE` writes a UTF-8 byte-order mark, which is what makes Excel
open accented characters correctly. `.csv2` writes semicolon-separated values
with a comma decimal mark, the convention Czech and German Excel expects.

## Reproducibility

`dr_check_code()` runs your scripts in a clean R subprocess and reports which
ones fail — it catches the script that only works because of an object left
over in your workspace. Guard the export at the end of a script so the check
does not itself write an export:

```r
if (!dr_is_rerun()) {
  save_outputs(output_dir, list_outputs, code_location = code_locations)
}
```

`setup_code()` writes a starter script already laid out this way.

## Coming from the standalone script

The API is deliberately close to the original. An old `list_outputs.rds` reads
back with `as_outputs()`, which also maps the former `gpars$ggplot` to
`gpars$sizes`:

```r
old <- readRDS("01_output/01_list_outputs.rds")
out <- as_outputs(old)
```

See `NEWS.md` for the full list of differences.
